Supplementary material

### 2017-07-05 Manuscript/ms.tex
Updates to intro and model structure

We know mosquitoes spread dengue, but mosquito abundance, as estimated from trap-based surveillance data, is not causally linked to human disease risk.

@@

We know mosquitoes spread dengue, but mosquito abundance, as estimated from trap-based surveillance data, is not causally linked to human disease risk.

@@

Dengue fever is a massive global public health burden, with millions of cases per year in Brazil alone (CITATION).  
Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
Though eradication of \textit{Aedes aegypti} is infeasible, effectively targeted control interventions may still substantially reduce disease risk.

Cities deploy mosquito control based on trap-based mosquito surveillance data.
For instance, the "MI-Dengue" system deployed throughout several cities in Brazil uses a grid of mosquito sticky-traps (MosquiTRAPs) for spatial prioritization of source reduction and larvicide application \cite{Eiras2009}.
Comparing cities with and without this system, \cite{Pepin2013} estimated that it prevented X cases and saved X dollars in economic losses from 20XX - 20XX.
Though this system has been effective, the mosquito indices and thresholds used to prioritize control efforts remain somewhat ad-hoc.
Using such surveillance data to efficiently target limited control resources requires a more detailed understanding of the relationship between the components of mosquito abundance measured by traps and disease risk.

Many of the attempts to establish this relationship statistically have relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
For example, \cite{Pepin2015} found that models that included MosquiTRAP surveillance data failed to predict weekly dengue cases any better than models that include lagged case data alone.
However, these models fail to account for the mechanistic, non-linear relationship between mosquito population dynamics and dengue epidemiology.
In particular, the cycle of transmission between humans and dengue is controlled not just by mosquito population size, but also by the emergence of new susceptible adults and the mortality rate.

In this paper, we seek to connect mosquito surveillance data with human case reports through the use of a Bayesian, mechanistic model.
We first propose a differential equation model for dengue epidemiology.
We then embed this model into a Bayesian statistical framework that allows us to estimate latent time series of mosquito demographic parameters from time series of mosquito trap counts and reported cases of dengue fever.
Lastly, we explore the ability of this fitted model to predict annual dengue outbreaks and its utility for targeting control efforts.

@@

Explicitly accounting for interactions between serotypes, even only two of the four, leads to a large and complex mechanistic model.
Moreover, since dengue case reports do not include information on serotype, we do not have enough information to inform the dynamics of individual serotypes.
As such, we simplify the model of \cite{Wearing2006} to an SEIRS framework, which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
In this framework, susceptible individuals ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
Following a latent period ($\frac{1}{\rho}$), exposed individuals become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
Infectious individuals recover at rate $\gamma$ and subsequenty remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.

Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
Total mosquito population size ($V_N$) is controlled by a periodically varying net emergence rate, $b_t$, and a stochastic, weather dependent death rate, $d_t$.
Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.

The complete differential equation model is then given as:
\begin{align} 
\frac{dS}{dt} &= bN - bS - \lambda \frac{V_{I}}{N} S + \delta R\\
\frac{dE}{dt} &= \lambda \frac{V_{I}}{N} S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
\frac{dV_S}{dt} & = b_t - \lambda \frac{I}{N} V_S - (d_t + \phi_q \tau_t) V_S \\
\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{vt} + d_t + \phi_q \tau_t)V_E\\
\frac{dV_I}{dt} &= \rho_{vt} V_E - (d_t + \phi_q \tau_t) V_I\\
\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
V_N &= V_S + V_E + V_I,
\end{align}
where
\begin{equation}
\log(b_t) = \alpha_{0} + \alpha_1 \sin(2\pi t / 52) + \alpha_2 \cos(2\pi t / 52),
\end{equation}
and
\begin{align}
\log(d_{t}) & = \beta_{0} + X\beta + \epsilon_t\\
\epsilon_t & \sim \text{Normal}(0, \sigma).
\end{align}

>>>>

Mosquito abundance alone is not a strong predictor of dengue fever risk, but provides information that, when coupled with case data, allows us to estimate mosquito demographic rates.

@@

Mosquitoes are dumb.

@@

Dengue fever is a massive global public health burden, with millions of cases per year in Brazil alone \cite{Bhatt2013}.  
Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
To be effective, control efforts should be guided by consistent, high resolution mosquito surveillance \cite{Morrison2008}.
The "MI-Dengue" system, implemented by the company Ecovec and deployed throughout cities in Brazil, uses a city-wide grid of mosquito sticky-traps (MosquiTRAPs) for spatial prioritization of source reduction and larvicide application \cite{Eiras2009}.
Comparing cities with and without this system, \cite{Pepin2013} estimated that it prevented 27,191 cases of dengue fever from 2007 - 2011.
Though this system has been effective, the mosquito indices and thresholds used to prioritize control efforts remain somewhat ad-hoc.
As noted in \cite{Morrison2008}, entomological indices are only relevant in the context of local epidemiology, and thus control policies will be most effective when they synthesize both entomological and clinical surveillance.

Many of the attempts to link these two sources of information have found a weak relationship between mosquito surveillance disease risk (CITATIONS).
These efforts have often relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
For example, \cite{Pepin2015} found that models that included MosquiTRAP surveillance data failed to predict weekly dengue cases any better than models that included lagged case data alone.
However, such tests fail to account for the mechanistic, non-linear relationship between mosquito population dynamics and dengue epidemiology.
The cycle of transmission between humans and dengue is controlled not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period \cite{Achee2015}.
In fact, the probability that an exposed mosquito will survive long enough to become infectious provides a more direct measure of transmission potential than mosquito abundance alone.

Adult surveillance data (e.g. time series of mosquito trap counts) is insufficient on its own to estimate mosquito survival rates and thus transmission potential.
However, we can borrow from the field of 'integrated population modeling' from wildlife ecology and use multiple sources of information to estimate mosquito demographic rates (CITATIONS).
Since mosquito survival is a critical bottleneck in the transmission process, it can be informed from the observed transmission dynamics (i.e. reports of human cases), given an appropriate underlying mechanistic model.

In this paper, we propose a mechanistic transmission model that connects mosquito abundance and survival with dengue epidemiology and human case reports.
We then embed this model into a Bayesian statistical framework that allows us to estimate latent time series of mosquito demographic parameters from time series of mosquito trap counts and reported cases of dengue fever.
Lastly, we explore the utility of these estimated transmission risks for guiding targeted control efforts relative to using mosquito abundance alone.

@@

Explicitly accounting for interactions between serotypes, even only two of the four, leads to a large and complex mechanistic model.
Moreover, since dengue case reports do not include information on serotype, we do not have enough information to inform the dynamics of individual serotypes.
As such, we simplify the model of \cite{Wearing2006} to an SEIRS framework, which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
In this framework, susceptible individuals ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
Following a latent period ($\frac{1}{\rho}$), exposed individuals become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
Infectious individuals recover at rate $\gamma$ and subsequently remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.

Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
Total mosquito population size ($V_N$) is controlled by stochastic, seasonally varying growth rate ($r_t$) and death rate ($d_t$).
Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.

The complete differential equation model is then given as:
\begin{align} 
\frac{dS}{dt} &= bN - bS - \lambda \frac{V_{I}}{N} S + \delta R\\
\frac{dE}{dt} &= \lambda \frac{V_{I}}{N} S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
\frac{dV_N}{dt} & = r_t V_N \\
\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{vt} + d_t + \phi_q \tau_t)V_E\\
\frac{dV_I}{dt} &= \rho_{vt} V_E - (d_t + \phi_q \tau_t) V_I\\
\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
V_S &= V_N - V_E - V_I,
\end{align}
where
\begin{align}
d_t & = \rho_{vt} \exp(-\nu_t)\\
\nu_{t} & = \alpha_{\nu} \nu_{t-1} + \theta_{\nu} (\nu_{t-52} - \alpha_{\nu} \nu_{t-53}) +\epsilon_{\nu t}\\
r_t &= \alpha_r r_{t-1} + \theta_r (r_{t-52} - \alpha_r r_{t-53}) + \epsilon_{rt},\\
\end{align}
and
\begin{align}
\epsilon_{rt} & \sim \text{Normal}(0, \sigma_r)\\
\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma_{\nu}).
\end{align}
